Title

### 2013-02-25 inc/extension/external_forces/external_forces.tex
improve formal external_forces section

Besides a set of initial conditions, the Analytic Gaussian model presented by Michalik and Sipe (see Section \ref{sec:ms_model}) is only useful for modeling free-space propagation.
This is because there is no allowance for external forces, such as those imparted by the ``optical'' elements of an electron microscope column.

@@

The internal Coulombic repulsion force enters the model in \ref{eq:potential_integral}.
Since both potentials and integrals are linear, one may add additional potentials so that it becomes
\begin{equation}
  V(\vec{r} - \vecprime{r}) \rightarrow V(\vec{r} - \vecprime{r}) + \sum_{k} V_{k}^{ext}(\vec{r})
\end{equation}
Following this new term through the derivation, we see that
\begin{equation}
  \begin{split}
    \Phi(\vec{r};t) = & \int d\vecprime{r} \left( V(\vec{r} - \vecprime{r}) + \sum_{k} V_{k}^{ext}(\vec{r}) \right) n(\vecprime{r}; t) \\
    = & \int d\vecprime{r} V(\vec{r} - \vecprime{r}) n(\vecprime{r}; t) + \sum_{k} \int d\vecprime{r} V_{k}^{ext}(\vec{r}) n(\vecprime{r}; t) \\
    \equiv & \Phi^{int}(\vec{r};t) + \sum_{k} \Phi_{k}^{ext}(\vec{r};t)
  \end{split}
\end{equation}
Therefore each external force contributes an additional term defined as
\begin{equation}
  \Phi^{ext}(\vec{r};t) \equiv \int d\vecprime{r} V^{ext}(\vec{r}) n(\vecprime{r}; t)
\end{equation}
These external terms are not self-influencing, and thus are not subject to the complexities of the internal Coulombic repulsion force. For clarity we now define $\Phi^{int}(\vec{r};t)$ as the old $\Phi(\vec{r};t)$, which results from the Coulomb force, while our new $K^{force}$ now includes these exteral terms.

This form of $\Phi(\vec{r};t)$ can now be substituted into the definition of $K^{force}$ matrix (\ref{eq:Kforce}), again in the basis defined in \ref{eq:coordinate_vector} 
\begin{equation}
  \begin{split}
    K^{force}_{ij} = & \frac{1}{N} \int u_i u_j \frac{\partial f}{\partial \vec{p}} \frac{\partial}{\partial \vec{r}} \left( \Phi(\vec{r};t) + \sum_{k} \Phi_{k}^{ext}(\vec{r};t) \right) d\vec{r} d\vec{p} \\
    \equiv & K^{int}_{ij} + \sum_{k} (K_{k}^{ext})_{ij}
  \end{split}
\end{equation}
Like the potential integrals, we are left with an additional matrix for each external force. Again, for clarity, we now define $K^{int}$ as the $K$ matrix resulting from the internal Coulombic repulsion.

@@

For constant static forces having no dependence in $ x $, $ y $, $ z $, or $ t $, one can readily show that there is no effect on the pulse propagation parameters $ \sigma_{ \alpha } $, $ \eta_{ \alpha } $, and $ \gamma_{ \alpha } $.
For example, consider a field consider a field $ \vec{E} = -E \hat{z} $ such that electrons are accelerated in the positive $ z $ direction by $ \vec{F}_{ext} = qE \hat{z} = q ( V / d ) \hat{z} $, so that from \ref{eq:Kint} we have
\begin{equation}
 \begin{split} 
  K^{ext}_{ij} = \frac{1} {N} \dfrac{qV} {d} \int{ u_{i} u_{j} \left( \frac{p_{z}} {\eta_{z}} - \frac{\gamma_{z} z} {\sigma_{z} \eta_{z}} \right) f d^{3}x   d^{3}p } ;
 \end{split}
\end{equation}
but it is clear that no function of this form multiplying $ f $, which is a Gaussian distribution, can avoid creating an odd integrand, and thus $ K_{ij}^{ext} = 0 $ for all $ i $ and $ j $ (or equivalently $ K^{ext} = \hat{0} $) meaning that this static field has no effect on the parameters of the pulse.
Of course, this is only true in the pulse frame, where there is no total momentum and thus $ \Delta E = \left( \Delta p \right)^{2} / 2 m_{e} $ is unchanged.
However, in the lab frame, where  $ \Delta E = p \Delta p / m_{e} $, there is an increase in the momentum spread, as the pulse now has the large momentum $p$ imparted by the accelerating field.

@@

For the case of a perfect lens, or in the longitudinal direction a perfect RF compression element, with a parabolic potential, the external force is linear with distance from the center of the electron pulse: 
\begin{equation} \label{eq:linear_force}
\vec{F}_{ext} = - M_{ \smallT } r \hat{\operatorname{r}} - M_{z} z \hat{\operatorname{z}} \text{ .}
\end{equation}
Here, $M_{ \smallT }$ and $M_{z}$ characterize the strength of a lens or a compressor, respectively, and are chosen to be positive for pulse focusing. Substitution into the integral equation for the $ K^{ext} $ matrix elements,
\begin{equation}
 \begin{split} 
  K^{ext}_{ij} & \equiv \dfrac{1} {N} \int u_{i} u_{j} ( \vec{\nabla}_{p} f ) \cdot \vec{F}_{ext} d^{3} x d^{3} p \\ & = \dfrac{1} {N} \int u_{i} u_{j} \biggl[ x \left( \dfrac{ p_{x} } { \eta_{ \smallT } } - \dfrac{ \gamma_{ \smallT } x } { \sigma_{ \smallT } \eta_{ \smallT } } \right) + y \left( \dfrac{ p_{y} } { \eta_{ \smallT } } - \dfrac{ \gamma_{ \smallT } y } { \sigma_{ \smallT } \eta_{ \smallT } } \right) \\ & \quad + z \left( \dfrac{ p_{z} } { \eta_{z} } - \dfrac{ \gamma_{z} z } { \sigma_{z} \eta_{z} } \right) \biggr] f d^{3} x d^{3} p ,
 \end{split}
\end{equation}
leads to the $6\times6$ matrix
\begin{equation}
K^{ext} = 
\begin{pmatrix}
\hat{k}^{lin}_{\smallT} & \hat{0} & \hat{0} \\
\hat{0} & \hat{k}^{lin}_{\smallT} & \hat{0} \\
\hat{0} & \hat{0} & \hat{k}^{lin}_{z}
\end{pmatrix} \text{ ,}
\end{equation}
where
\begin{equation}
\hat{k}^{lin}_{\alpha} =
\begin{pmatrix}
0 & M_{\alpha} \sigma_{\alpha} \\
M_{\alpha} \sigma_{\alpha} & 2 M_{\alpha} \gamma_{\alpha} \text{.}
\end{pmatrix} \text{ .}
\end{equation}
which by \ref{eq:dainvdt} yields an additional term $M_{\alpha} \sigma_{\alpha}$ which is added to the original $\dfrac{d \gamma_{\alpha}}{d t}$ equation.

>>>>

Simulation of electron pulse propagation in UEM and modern UED apparatus\cite{siwick_ultrafast_2002,oudheusden_electron_2007,reed_femtosecond_2006} requires the ability to model electron-optical elements designed to control the three-dimensional space-charge and dispersion induced broadening of the pulse, and hence deliver electron pulses of the desired quality to the specimen. 
The Analytic Gaussian model as presented by Michalik and Sipe (see Section \ref{sec:ms_model}) is only useful for modeling free-space propagation.
This is because there is no allowance for external forces, such as those imparted by the ``optical'' elements of an electron microscope column.
In this section, I show that the AG model can be extended to include external forces on the pulse, such as the influence of magnetic electron lenses and RF pulse compression cavities.\cite{oudheusden_electron_2007,veisz_hybrid_2007}

@@

The internal Coulombic repulsion force enters the model in \ref{eq:potential_integral}.
Since both potentials and integrals are linear, one may add additional potentials so that it becomes
\begin{equation}
  V(\vec{r} - \vecprime{r}) \rightarrow V^{int}(\vec{r} - \vecprime{r}) + \sum_{k} V_{k}^{ext}(\vec{r})
\end{equation}
In doing so, we rename the old potential, the internal repulsive Coulomb force, $V^{int}$.
Following these new terms through the derivation, we see that
\begin{equation}
  \begin{split}
    \Phi(\vec{r};t) = & \int d\vecprime{r} \left( V^{int}(\vec{r} - \vecprime{r}) + \sum_{k} V_{k}^{ext}(\vec{r}) \right) n(\vecprime{r}; t) \\
    = & \int d\vecprime{r} V^{int}(\vec{r} - \vecprime{r}) n(\vecprime{r}; t) + \sum_{k} \int d\vecprime{r} V_{k}^{ext}(\vec{r}) n(\vecprime{r}; t) \\
    \equiv & \Phi^{int}(\vec{r};t) + \sum_{k} \Phi_{k}^{ext}(\vec{r};t)
  \end{split}
\end{equation}
Therefore each external force contributes an additional term defined as
\begin{equation}
  \Phi^{ext}(\vec{r};t) \equiv \int d\vecprime{r} V^{ext}(\vec{r}) n(\vecprime{r}; t)
\end{equation}
These external terms are not self-influencing, and thus are not subject to the complexities of the internal Coulombic repulsion force.
For consistency we have now defined $\Phi^{int}(\vec{r};t)$ as the old $\Phi(\vec{r};t)$, which results from the Coulomb force.

This new form of $\Phi(\vec{r};t)$ can be substituted into the definition of $K^{force}$ matrix (\ref{eq:Kforce}), again in the basis defined in \ref{eq:coordinate_vector} 
\begin{equation}
  \begin{split}
    K^{force}_{ij} = & \frac{1}{N} \int u_i u_j \frac{\partial f}{\partial \vec{p}} \frac{\partial}{\partial \vec{r}} \left( \Phi(\vec{r};t) + \sum_{k} \Phi_{k}^{ext}(\vec{r};t) \right) d\vec{r} d\vec{p} \\
    \equiv & K^{int}_{ij} + \sum_{k} (K_{k}^{ext})_{ij}
  \end{split}
\end{equation}
Like the potential integrals, we are left with an additional matrix for each external force. Again, for consistency, we have defined $K^{int}$ as the $K$ matrix resulting from the internal Coulombic repulsion.

@@

For constant static forces having no dependence in $ x $, $ y $, $ z $, or $ t $, one can readily show that there is no effect on the pulse propagation parameters $ \sigma_{ \alpha } $, $ \eta_{ \alpha } $, and $ \gamma_{ \alpha } $.
For example, consider a field consider a field $ \vec{E} = -E \hat{z} $ such that electrons are accelerated in the positive $ z $ direction by $ \vec{F}_{ext} = qE \hat{z} = q ( V / d ) \hat{z} $, so that from \ref{eq:Kint} we have
\begin{equation}
 \begin{split} 
  K^{ext}_{ij} = \frac{1} {N} \dfrac{qV} {d} \int{ u_{i} u_{j} \left( \frac{p_{z}} {\eta_{z}} - \frac{\gamma_{z} z} {\sigma_{z} \eta_{z}} \right) f d^{3}x   d^{3}p } ;
 \end{split}
\end{equation}
but it is clear that no function of this form multiplying $ f $, which is a Gaussian distribution, can avoid creating an odd integrand, and thus $ K_{ij}^{ext} = 0 $ for all $ i $ and $ j $ (or equivalently $ K^{ext} = \hat{0} $) meaning that this static field has no effect on the parameters of the pulse.
Of course, this is only true in the pulse frame, the frame of the simulation, where there is no total momentum and thus $ \Delta E = \left( \Delta p \right)^{2} / 2 m_{e} $ is unchanged.
In the lab frame however, where  $ \Delta E = p \Delta p / m_{e} $, there is an increase in the momentum spread, as the pulse now has the large momentum $p$ imparted by the accelerating field.

@@

For the case of a perfect lens, or in the longitudinal direction a perfect RF compression element, with a parabolic potential, the external force is linear with distance from the center of the electron pulse: 
\begin{equation} \label{eq:linear_force}
\vec{F}_{ext} = - M_{ \smallT } r \hat{\operatorname{r}} - M_{z} z \hat{\operatorname{z}} \text{ .}
\end{equation}
Here, $M_{ \smallT }$ and $M_{z}$ characterize the strength of a lens or a compressor, respectively, and are chosen to be positive for focusing/compression elements.
Substitution into the integral equation for the $ K^{ext} $ matrix elements,
\begin{equation}
 \begin{split} 
  K^{ext}_{ij} & \equiv \dfrac{1} {N} \int u_{i} u_{j} ( \vec{\nabla}_{p} f ) \cdot \vec{F}_{ext} d^{3} x d^{3} p \\ & = \dfrac{1} {N} \int u_{i} u_{j} \biggl[ x \left( \dfrac{ p_{x} } { \eta_{ \smallT } } - \dfrac{ \gamma_{ \smallT } x } { \sigma_{ \smallT } \eta_{ \smallT } } \right) + y \left( \dfrac{ p_{y} } { \eta_{ \smallT } } - \dfrac{ \gamma_{ \smallT } y } { \sigma_{ \smallT } \eta_{ \smallT } } \right) \\ & \quad + z \left( \dfrac{ p_{z} } { \eta_{z} } - \dfrac{ \gamma_{z} z } { \sigma_{z} \eta_{z} } \right) \biggr] f d^{3} x d^{3} p ,
 \end{split}
\end{equation}
leads to the $6\times6$ matrix
\begin{equation}
K^{ext} = 
\begin{pmatrix}
\hat{k}^{lin}_{\smallT} & \hat{0} & \hat{0} \\
\hat{0} & \hat{k}^{lin}_{\smallT} & \hat{0} \\
\hat{0} & \hat{0} & \hat{k}^{lin}_{z}
\end{pmatrix} \text{ ,}
\end{equation}
where
\begin{equation}
\hat{k}^{lin}_{\alpha} =
\begin{pmatrix}
0 & M_{\alpha} \sigma_{\alpha} \\
M_{\alpha} \sigma_{\alpha} & 2 M_{\alpha} \gamma_{\alpha} \text{.}
\end{pmatrix} \text{ .}
\end{equation}
which by \ref{eq:dainvdt} yields an additional term $M_{\alpha} \sigma_{\alpha}$ which is added to the original $\dfrac{d \gamma_{\alpha}}{d t}$ equation.
This derivation has been for a single force, in practice however there may be multiple forces, therefore \ref{eq:ag_big_gamma_gamma} may be restated as
\begin{equation}
  \frac{d\gamma_{\alpha}}{dt} = \frac{ \Gamma_{\alpha}^2(0) + \gamma_{\alpha}^2 }{\sigma_{\alpha} m}
    + \frac{N e^2}{4\pi\varepsilon_0} \frac{1}{6 \sqrt{\sigma_{\alpha}\pi}} L_{\alpha}(\xi) + \sum M_{\alpha} \sigma_{\alpha}
\end{equation}

Since these terms are only added to $\dfrac{d \gamma_{\alpha}}{d t}$, the analysis of Liouville's Theorem, shown in Section \ref{sec:liouville}, is unaffected.
The extended AG model still explicitly satisfies Liouvilles's Theorem.
